---
title: "Check alignment quality by STAR of batch 2"
author: "Caroline Lohoff"
date: "December 13 2021"
output:
  BiocStyle::html_document:
    toc_float: true
---

# Set options and load packages and data

Set global options
```{r setup, include=FALSE, cache = FALSE}
#knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_knit$set(root.dir = "~/Documents/R/drugseq_test")
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Load packages
```{r}
library(tidyverse)
```

Load patient annotation information into data frame
```{r}
patAnno <- readxl::read_xlsx("./data/RNAseq_samples_pilot_batch1old_batch1new_batch2.xlsx") %>%
  mutate(fileName = str_replace(fileName, ".txt.gz","")) %>%
  mutate(Replicate = as.factor(Replicate)) %>%
  select(-sequence) %>%
  mutate(cell = ifelse(patID == "Jurkat","Jurkat","Primary")) %>%
  select(fileName, expBatch, patID, sampleID, batch, cell)
```


# Quality information from STAR alignment (genome mapping)

Define function
```{r}
processLog <- function(fileName) {
    outTab <- tibble(allLine = read_lines(fileName)) %>%
      separate(allLine, c("entry","value"),"[|]", remove = FALSE) %>%
      mutate(value = str_remove_all(value, "[\t%]"),
             entry = str_trim(entry,"both")) %>%
      mutate(value = as.numeric(value)) %>%
      filter(!is.na(value)) %>% select(-allLine)
    
    outTab
}
```

Load bam files from pilot batch, batch 1 old, batch 1 new, and batch 2
```{r, warning=FALSE, error=FALSE}
allFiles <- list.files("~/Documents/drugSeq/bam/pilot_batch1old_batch1new_batch2/",recursive = TRUE, full.names = TRUE)
allRecord <- lapply(allFiles, function(fileName){
  baseName <- basename(fileName)
  processLog(fileName) %>%
    mutate(filename = str_remove(baseName, "Log.final.out"))
}) %>% bind_rows()
```

Check that all samples have patient annotations
```{r}
all(allRecord$filename %in% patAnno$fileName)
all(patAnno$fileName %in% allRecord$filename) 

allRecord <- left_join(allRecord, patAnno, by = c(filename = "fileName")) %>% arrange(entry)
```

Create multiple box plots
```{r, fig.width=15, fig.height=25}
diagBoxPlots <- lapply(unique(allRecord$entry), function(n) {
  plotTab <- filter(allRecord, entry == n)
  ggplot(plotTab, aes(x=batch, y=value)) +
    geom_boxplot(outlier.shape = NA) +
    ggbeeswarm::geom_quasirandom(aes(col=cell), groupOnX=TRUE) +
    theme_bw() + theme(legend.position = "none") +
    ggtitle(n)
})
cowplot::plot_grid(plotlist = diagBoxPlots, ncol=3)
```


# Quality information from HTseq (gene counting)

Load data
```{r}
htTab <- read_tsv("./output/htseq_process_batch2_20211213.tsv")
htTab <- htTab %>% mutate(batch = patAnno[match(fileName, patAnno$fileName),]$batch,
                          cell = patAnno[match(fileName, patAnno$fileName),]$cell) %>%
  mutate(percent = percent*100)
```

Create multiple box plots
```{r, fig.width=10, fig.height=8}
diagBoxPlots <- lapply(unique(htTab$feature), function(n) {
  plotTab <- filter(htTab, feature == n)
  ggplot(plotTab, aes(x=batch, y=percent)) +
    geom_boxplot(outlier.shape = NA) +
    ggbeeswarm::geom_quasirandom(aes(col = cell), groupOnX=TRUE) +
    ylab("Percentage") +
    theme_bw() + theme(legend.position = "none") +
    ggtitle(n)
})
cowplot::plot_grid(plotlist = diagBoxPlots, ncol=3)
```


# Combine and save Star and HTseq
```{r}
htTabReformated <- select(htTab, feature, count, percent, fileName) %>%
  pivot_longer(c(count, percent), names_to = "type", values_to = "value") %>%
  mutate(entry = ifelse(type == "count", feature, paste0(feature," %"))) %>%
  select(value, entry, fileName) %>%
  left_join(select(patAnno, fileName, expBatch, patID, sampleID, batch), by = "fileName") %>%
  mutate(source = "HTseq") %>%
  dplyr::rename(filename  = fileName)

allRecordOut <- bind_rows(allRecord %>% mutate(source = "STAR"),
                          htTabReformated)

write_csv2(allRecordOut, file = "./output/STAR_HTseq_batch2_summary_20211213.csv")
```


# Salmon results

## Load data

Define function for loading data
```{r}
load_salmon <- function(dir_salmon){
  lapply(dir_salmon, function(n) {
    filename <- basename(n)
    allLines <- read_lines(paste0(n,"/logs/salmon_quant.log"))
    allLines <- allLines[str_detect(allLines, "Mapping rate =")]
    mapRate <- str_split(allLines, "=")[[1]][2] %>% str_remove_all("[ %]") %>% as.numeric()
  tibble(filename = filename, value = mapRate)
  }) %>% bind_rows() %>%
    left_join(patAnno, by = c(filename = "fileName"))
}
```

Pilot batch
```{r}
allDir <- list.dirs("~/Documents/drugSeq/salmon/pilot/", recursive = FALSE)
salmonOut.pilot <- load_salmon(allDir)
```

Batch 1 old
```{r}
allDir <- list.dirs("~/Documents/drugSeq/salmon/batch1/", recursive = FALSE)
salmonOut.batch1old <- load_salmon(allDir)
```

Batch 1 new
```{r}
allDir <- list.dirs("~/Documents/drugSeq/salmon/batch1_new/", recursive = FALSE)
salmonOut.batch1new <- load_salmon(allDir)
```

Batch 2
```{r}
allDir <- list.dirs("~/Documents/drugSeq/salmon/batch2/", recursive = FALSE)
salmonOut.batch2 <- load_salmon(allDir)
```

## Plot mapping rate
```{r, fig.height=4, fig.width=6}
salmonOut <- bind_rows(salmonOut.pilot, salmonOut.batch1old,
                       salmonOut.batch1new, salmonOut.batch2)
ggplot(salmonOut, aes(x=batch, y=value)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(aes(col = cell)) +
  ylab("Percentage")  +
  theme_bw() + theme(legend.position = "none") +
  ggtitle("Salmon mapping rate")
```


# Correlation quality features with cell viability (STAR & HTseq)

Load data
```{r}
load("./data/patmeta_210324.RData")

allRecordOut <- read_csv2("./output/STAR_HTseq_batch2_summary_20211213.csv") %>%
  filter(str_detect(entry,"%"), batch == "batch1") %>%
  select(entry, value, filename)

patAnno <- readxl::read_xlsx("./data/RNAseq_samples_pilot_batch1old_batch1new_batch2.xlsx") %>%
  mutate(fileName = str_replace(fileName, ".txt.gz","")) %>%
  mutate(diagnosis = patMeta[match(patID, patMeta$Patient.ID),]$diagnosis) %>%
  mutate(diagnosis = ifelse(patID == "Jurkat","Jurkat",diagnosis))

patTest <- patAnno %>%
  select(fileName, FSC.SSC.d0, FSC.SSC, FSC.SSC.norm) %>%
  pivot_longer(-fileName, names_to = "metaVar", values_to = "metaValue")
```

Show results
```{r}
testTab <- full_join(allRecordOut, patTest, by = c(filename= "fileName")) %>%
  filter(!is.na(value), !is.na(metaValue))
  
resTab <- group_by(testTab, entry, metaVar) %>% nest() %>%
  mutate(m=map(data, ~cor.test(~value + metaValue, .,
                              method = "spearman", use = "pairwise.complete.obs"))) %>%
  mutate(res = map(m, broom::tidy)) %>% 
  unnest(res) %>%
  select(entry, metaVar, p.value, estimate) %>%
  arrange(p.value)

resTab %>% mutate_if(is.numeric, formatC) %>%
  DT::datatable()
```


# Top 12 associations as scatter plot

All samples
```{r, fig.height=9, fig.width=12, warning=FALSE, message=FALSE}
pList <- lapply(seq(12), function(i) {
  rec <- resTab[i,]
  plotTab <- filter(testTab, entry == rec$entry, metaVar == rec$metaVar) %>%
    mutate(diagnosis = patAnno[match(filename, patAnno$fileName),]$diagnosis)
  ggplot(plotTab, aes(x=metaValue, y=value)) +
    geom_point(aes(col = diagnosis)) + geom_smooth(method = "lm") +
    theme_bw() +
    xlab(rec$metaVar) + ylab(rec$entry)
})

cowplot::plot_grid(plotlist = pList, ncol=3)
```

Only CLL samples


# Session Info Details
```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```

