---
title: "Plot for patient overview (mutations)"
author: "Caroline Lohoff"
date: "June 6 2022"
output:
  BiocStyle::html_document:
    toc_float: true
---

# Description
The aim of this script is to generate a plot which presents an overview of the patients from which samples were extretced and used for the CLL data set.
This plots contains information about the mutations and clinical background of the patients.

# Load libraries and data sets

Set global options
```{r setup, include=FALSE, cache = FALSE}
knitr::opts_knit$set(root.dir = "~/Documents/R/drugseq_test")
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Load libraries
```{r load-libs}
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(pheatmap)
  library(RColorBrewer)
})
```

Set global ggplot variables
```{r set-ggplot}
theme_set(theme_bw() + theme(axis.text = element_text(size=12), 
                             axis.title = element_text(size=14),
                             legend.title = element_text(size=14),
                             legend.text = element_text(size=12),
                             plot.title = element_text(size=16,
                                                       hjust=0.5,face="bold")))
```


# Load data and add annotations

Load datasets
```{r}
# Load meta data including genotyping info
load("./data/patmeta_210324.RData")
patMeta$pretreated <- patMeta$treatment

# Load survival data
load("./data/survival_190516.RData")
names(survT)[names(survT) == "treatment"] <- "treatment_bin"

# Load dds data set containing 1106 samples and count matrix
load("./output/ddsDrug_htseq_batchPilot1c234_20220426.RData")

# Add columns to DESeq object from survival data
sur_vec <- c("sampleDate", "LKA", "OS", "died", "diagDate", "TFT", "TTT",
             "treatment_bin", "firstTreatDate", "lastKnownUntreated", "treatedAfter")
ddsDrug@colData@listData <- lapply(sur_vec, function(n) {
  ddsDrug[[n]] <- survT[match(ddsDrug$sampleID, survT$sampleID), n]
}) %>% bind_cols() %>% sjmisc::add_columns(ddsDrug@colData@listData)

# Add columns to DESeq object from patient meta data
col_vec <- c("diagnosis", "gender", "pretreated", "Methylation_Cluster", "trisomy12", "NOTCH1", "TP53",
             "ATM", "FBXW7", "MED12", "BRAF", "SF3B1", "del11q", "del17p", "del13q", "del8p", "gain2p",
             "gain8q", "del6q", "gain3q", "del9p", "del15q", "del6p", "MYD88", "KRAS", "EGR2", "XPO1")
ddsDrug@colData@listData <- lapply(col_vec, function(n) {
  ddsDrug[[n]] <- patMeta[match(ddsDrug$patID, patMeta$Patient.ID), n]
}) %>% bind_cols() %>% sjmisc::add_columns(ddsDrug@colData@listData)

names(ddsDrug@colData@listData)[names(ddsDrug@colData@listData) == "Methylation_Cluster"] <- "Methylation"
ddsDrug$IGHVpat <- patMeta[match(ddsDrug$patID, patMeta$Patient.ID),]$IGHV.status
ddsDrug$IGHV <- ifelse(is.na(ddsDrug$IGHV), ddsDrug$IGHVpat, ddsDrug$IGHV)
ddsDrug$IGHVpat <- NULL
ddsDrug$IGHV[ddsDrug$batch == "pilot" & ddsDrug$patID %in% c("PID1095", "PID1222")] <- "U"
ddsDrug$IGHV[ddsDrug$batch == "pilot" & ddsDrug$patID %in% c("PID1177", "PID1210")] <- "M"
ddsDrug$IGHV[ddsDrug$IGHV == "U"] <- 0
ddsDrug$IGHV[ddsDrug$IGHV == "M"] <- 1

# Add meta data to the samples from the pilot batch
ddsDrug$diagnosis[ddsDrug$batch == "pilot"] <- "CLL" 
ddsDrug$trisomy12[ddsDrug$batch == "pilot"] <- 0
ddsDrug$gender[ddsDrug$batch == "pilot" & ddsDrug$patID %in% c("PID1095", "PID1222")] <- "f"
ddsDrug$gender[ddsDrug$batch == "pilot" & ddsDrug$patID %in% c("PID1177", "PID1210")] <- "m"

# Add age of patients on day of sample extraction
ddsDrug$age <- ageTab[match(ddsDrug$sampleID, ageTab$sampleID),]$age

# Remove replicates
ddsDrug <- ddsDrug[ ,ddsDrug$Replicate == 1]

# Retrieve patient annotations from dds object as data frame
patAnno <- colData(ddsDrug) %>% as_tibble()

dim(ddsDrug)
```

# Barplots 

## Number of samples per treatment

```{r, fig.width=8, fig.height=5}
color_vec <- colorRampPalette(brewer.pal(1,"Pastel2"))(12)

patAnno %>%
  ggplot(aes(x=treatment, fill=treatment)) +
  geom_bar() +
  geom_text(stat = "count", aes(label=..count..),
            vjust=1.3, size=4, color="#323232") +
  theme(axis.text.x = element_text(angle = 45, vjust=0.9, hjust=0.9)) +
  ylab("Number of samples") +
  theme(axis.title.x = element_blank(),
        legend.position="none") +
  scale_fill_manual(values=color_vec)
```

## Pathology classification of samples

```{r, fig.width=4, fig.height=5}
# Remove baseline samples (Jurkat cell line)
ddsDrug <- ddsDrug[ ,ddsDrug$treatment != "Baseline"]
patAnno <- colData(ddsDrug) %>% as_tibble()
dim(ddsDrug)

# Keep every patient only one time
patAnno.sub <- patAnno[!duplicated(patAnno$patID), ]
length(patAnno.sub$patID)

color_vec <- colorRampPalette(brewer.pal(10,"Set3"))(5)

patAnno.sub %>%
  ggplot(aes(x=diagnosis, fill=diagnosis)) +
    geom_bar() +
  geom_text(stat = "count", aes(label=..count..),
            vjust=-0.2,size=4,color="#323232") +
    theme(axis.text.x = element_text(angle = 45, vjust=0.9, hjust=0.9)) +
    ylab("Number of patients") +
    theme(axis.title.x = element_blank(),
        legend.position="none") +
    scale_fill_manual(values=color_vec)
```


# Genetic landscape of CLL cohort

Filter data set by diagnosis
```{r}
ddsDrug <- ddsDrug[ ,ddsDrug$diagnosis %in% "CLL"]
patAnno <- colData(ddsDrug) %>% as_tibble()
dim(ddsDrug)
length(unique(patAnno$patID))
```

Extract expression matrix for DMSO
```{r}
exprMat <- counts(ddsDrug)

patAnno_DMSO <- patAnno[patAnno$treatment == "DMSO", ]
patAnno_DMSOnames <- patAnno_DMSO %>%
  column_to_rownames(var="ID")
exprMat_DMSO <- exprMat[,colnames(exprMat) %in% patAnno_DMSO$ID]
length(colnames(exprMat_DMSO))
```

```{r}
#Prepare data frame
annoCol <- patAnno_DMSO %>%
  select(ID, pretreated, age, gender, batch, IGHV, TP53, ATM,
         SF3B1, NOTCH1, MYD88, BRAF, KRAS, EGR2, MED12, FBXW7, XPO1,
         trisomy12, del17p, del13q, del11q, del8p, gain2p,
         gain8q, del6q, gain3q, del9p, del15q, del6p) %>%
  column_to_rownames(var="ID")

# Convert pretreated column to character
annoCol$pretreated[is.na(annoCol$pretreated)] <- 2
annoCol$pretreated <- as.character(annoCol$pretreated)
annoCol$pretreated[annoCol$pretreated == "0"] <- "No" 
annoCol$pretreated[annoCol$pretreated == "1"] <- "Yes" 
annoCol$pretreated[annoCol$pretreated == "2"] <- "na"

# Sort data frame by IGHV mutation status
annoCol <- annoCol[order(annoCol$IGHV, annoCol$TP53,
                         annoCol$del13q, decreasing = TRUE),]

# Subset data frame for header
annoColHeader <- annoCol %>%
  select(pretreated, age, gender, batch)

# Subset data frame for matrix
annoColMat <- annoCol %>%
  select(-batch, -gender, -pretreated, -age)

# Convert mutation columns from factor to numeric
annoColMat[is.na(annoColMat)] <- 2.0

ind <- sapply(annoColMat, is.factor)
annoColMat[ind] <- lapply(annoColMat[ind],
                          function(x) as.numeric(as.character(x)))
annoColMat$IGHV <- as.numeric(annoColMat$IGHV)

# Transpose and create matrix
mat <- t(annoColMat)

# Settings
breaks <- c(0,0.5,1)
colors <- c("white","black", "gray")
rowLabs <- rownames(mat)
```

```{r}
color_anno = list(
  batch = c(pilot="#F8F417", batch1="#B5E222", batch2="#76EE3D", batch3="#07D256", batch4="#0A9C43"),
  gender = c(f="#ECB3EE", m="#A201B5"),
  pretreated = c(Yes="#EFD08C", No="#8D6100", na="#A3884D"))
```

Plot heatmap
```{r, fig.width=10, fig.height=8}
plotGeneHeatmap <- pheatmap(mat, cluster_cols = FALSE, cluster_rows = FALSE, 
                            annotation_col = annoColHeader, annotation_colors=color_anno,
                            border_color = "black", treeheight_col=0,
         breaks = breaks, color = colors, show_colnames = T, fontsize=13,
         labels_col = patAnno_DMSOnames[colnames(mat),]$patID,
         labels_row = parse(text = rowLabs), fontsize_col=5,
         fontsize_row=10, legend = FALSE, silent = TRUE)$gtable
cowplot::plot_grid(plotGeneHeatmap)
```

As additional information survival data such as overall survival, died or time to treatment could be displayed in the heatmap.

# Session Info Details
```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```
